---
title: "Measles susceptibility in Edinburgh"
author: "Group 12"
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(readr)
library(ggplot2)
library(ggfortify)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
```

# Introduction {#sec-intro}

Data on Measles susceptibility in primary school children, between 1998 and 2012, in the 101 intermediate zones(IZ) in Edinburgh has been recorded. The IZ are small geographical areas that contain, on average, 4000 residents.

The data set DAProject16.csv contains the following data:

-   Y- The number of pre-school children susceptible to measles in a given IZ

-   N- The total number of pre-school children in a given IZ

-   Year- The year the data was collected

We are interested in whether Edinburgh exhibited a change in measles susceptibility following the retraction of the Wakefield article. We are also interested in if the change, if any, in measles susceptibility occurred in 2004 alongside the articles retraction.

# Exploratory Analysis {#sec-EA}

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-sum
#| tbl-cap: Mean,median and standard deviation of proportion by year.


measles<- read_csv("DAProject16.csv")

prop <- round((measles$Y/measles$N)*100, digits = 2)

measles <- transform(measles, prop = prop)

summary <- measles |>
  group_by(Year) |>
  summarise("Prop.Mean" = mean(prop),
            "Prop.Median" = median(prop),
            "Prop.sd" = sd(prop))

summary |>
  gt() |>
  fmt_number(columns = c(Prop.Mean,Prop.Median,Prop.sd),
             decimals = 2) |>
  tab_spanner(label = "Proportion",
              columns = c(Prop.Mean,Prop.Median,Prop.sd)) |>
  cols_label(
    Prop.Mean = html("Mean"),
    Prop.Median = html("Median"),
    Prop.sd = html("Std.Dev"))
```

@tbl-sum shows that the mean and median proportion of primary children susceptible to measles decreased after 2004. This could be attributed to the fact that article from Wakefield et al that linked the MMR vaccine with an increased risk of autism was partially retracted in 2004. @tbl-sum shows that the standard deviation of the proportion grew slightly after 2004 but decreased dramatically alongside the mean and median between 2008 and 2010.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-box
#| fig-cap: Proportion of primary children susceptible to measles by year. 

ggplot(data = measles, mapping = aes(x = factor(Year), y = prop)) +
  geom_boxplot() + labs(x = "Year", y = "Proportion(in %)")

```

@fig-box shows that the median of the box plots for 2008 and 2010 are much lower than the box plots of the other years.

# Formal Analysis  {#sec-FA}

We fit a normal linear model with...

# Conclusions {#sec-con}
